---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load 1,000 rows of data with the ages of people from 18-80

```{r}
set.seed(4321)

population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)
```


### Task 3: Sample 100 rows of data from the dataset

```{r}
sample100 <- population |>
  sample_n(10)
```


### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample100 |>
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )
```


### Task 5: Calculate the mean age and standard deviation of the population

```{r}
population |>
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )
```

### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The means and standard deviations are very similar. When using a sample size of 100, my mean age is 47.9 and my standard deviation is 17.6. When surveying the whole population, my mean age is 49.2 and my standard deviation is 18.4. When I changed my sample size from 100 to 10, my mean was 46.7 and my standard deviation was 16.5--a much larger difference than before. It seems as though the smaller my sample size is, the more inaccurate my results are--compared to if I polled the entire population. Overall though, each sampling size is within the same ballpark of each other.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
