---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

###The categories include a variety of incidents, ranging from serious crimes like homicide and sex offense to misdemeanors like theft. All of these categories are valuable from a reporting standpoint, as even  common crimes like theft and stolen auto vehicles can tell a larger story about a county. However, serious crimes like homicide, sex offense, and assault shootings hold significant weight in news because of their emotional appeal and fear factor--technically making them "important" from a reporting standpoint. Additionally, some of the categories make summarizing data difficult. For example, there's three categories for robbery: commercial, residential, vehicle, and other. Without clarification, your readers might be confused on what qualifies in each category or the sum of them together.

### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> mutate(total_incidents = rowSums(across(Jan:Dec)),
         average_incidents = total_incidents / 12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

###Minor crimes like car accidents, theft, and stolen auto vehicles have the highest average number of incidents per month. However, more serious crimes like homicide, assault shootings, and sexual offenses have the lowest average number of incidents per month. I think this is pretty on par with a lot of other urban counties, because densely-populated areas experience a high volume of non-violent crimes. I don't have many questions about the other categories, other than what "robbery other" means. 

### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> mutate(june_change = (Jun - average_incidents) / average_incidents * 100, july_change = (Jul - average_incidents) / average_incidents * 100, august_change = (Aug - average_incidents) / average_incidents * 100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

###Based on the percentage changes for June, July, and August, I don't think the theory about crime and summer months is sound. The June percentage changes are mostly positive--though there are a few categories that declined--showing that there is a slight increase in crime in June. However, in July and August, there is a mixture of categories declining and increasing. I don't think there's enough evidence for that theory--although there is an increase in homicides and stolen auto vehicles.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

###Lede: Assault shootings in Prince George's county saw a 34% increase in July compared to monthly averages.
###Nutgraf: Throughout the year, Prince George's county experienced 188 assault shootings, averaging nearly 16 incidents per month. The number of assault shootings rose to 21 in the month of July, a 34% increase compared to the monthly average. The only other months that saw this significant of an increase was August and October.


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
