---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}
ces <- ces |>
  mutate(evangelical = trad %in% c("White Evangelical"))

population_proportion <- mean(ces$evangelical)
```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

The population proportion for White Evangelicals is .104 (or 10.4%)--a far smaller number than the unaffiliated proportion of .36 (or 36%).

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42)
sample_50 <- ces |> sample_n(50)
mean(sample_50$evangelical)
```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

My sample estimate is .16, somewhat far off from the population proportion of .104. Similar to the unaffiliated group, a smaller sample size of 50 generally yields a less accurate representation of the entire dataset.

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}
set.seed(1966)
sample_500 <- ces |> sample_n(500)
mean(sample_500$evangelical)
```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

Just like in class, the larger sample size did improve the estimate. .09 is much closer to the population proportion of .104, suggesting that a larger sample size generally increases accuracy within polling.

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share White Evangelicals",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?

The shape of the n=50 histogram is somewhat similar to the unaffiliated histogram from class, though it's less symmetrical and slightly skewed to the right, with most samples clustered around .10. White Evangelicals only make up about 10% of the population, meaning the sample proportions are closer to zero. The data has more room to spread towards higher values, creating a longer right tail.

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

Similar to the polling activity in class, the n=500 histogram's distribution has a much narrower spread. Most estimates cluster closer to the population proportion of .104. However, the decrease in spread is smaller than what I saw in class, since the estimate for White Evangelicals were less spread out to begin with.

**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

Based on my simulation, there would be roughly 20 White Evangelical respondents. However, with a sample size of only 200 people, making estimates about a smaller subgroup would likely be less reliable--as only about 20 respondent belong to the subgroup.

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

As we covered throughout this assignment, estimates for smaller subgroups--like White Evangelicals--tend to be less reliable. Before reporting that number, I'd ask the campaign a few questions. First, did the poll survey 400 people overall or 400 White Evangelicals? How were these participants selected? Also, since the poll was shared by a campaign, was the way it was conducted, or the wording of its questions, biased? I'd also be curious what the margin of error for White Evangelicals was, compared to the full sample.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
